{TestIdeals} package for \emph{Macaulay2}

### 2019-04-24 Fsing/TestIdealsPaper/TestIdeals.tex
Renamed HSLGModule to FPureSubmodule.

Again, consider the maps $T^e : \omega_{R^{1/p^e}} \to \omega_R$ we have considered throughout this section.  It is a theorem of Hartshorne-Speiser, Lyubeznik, and Gabber \cite{HartshorneSpeiserLocalCohomologyInCharacteristicP,LyubeznikFModulesApplicationsToLocalCohomology,Gabber.tStruc} that the descending images
\begin{equation}\label{eq: descending images}
\omega_R \supseteq T(\omega_{R^{1/p}}) \supseteq \dots \supseteq T^e(\omega_{R^{1/p^e}}) \supseteq T^{e+1}(\omega_{R^{1/p^{e+1}}}) \supseteq \cdots
\end{equation}
stabilize for $e \gg 0$.
The function \texttt{HSLGModule} computes the stable submodule in this chain, called the \emph{HSLG module}, and returns a list containing the following items:
\begin{enumerate}[(1)]
   \item The HSLG module;
   \item The canonical module, embedded (non-uniquely) as an ideal;
   \item The element $u$ representing the map on the canonical module (see
   \autoref{subsec.DualToFrobeniusOnQuotientRings});
   \item The value of $e > 0$ at which the descending images in \eqref{eq: descending images} stabilize.  This is sometimes also called the HSLG-number of the canonical module as a Cartier module.
\end{enumerate}

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/3[x,y,z]/ideal( x^3 + y^4 + z^5 );
i2 : L = HSLGModule R;
i3 : L#0
                          2        2   3
o3 = ideal (0, y*z, x*z, y , x*y, x , z )
o3 : Ideal of R
i4 : L#1
o4 = ideal 1
o4 : Ideal of R
i5 : L#3
o5 = 1
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/2[x,y,z];
i2 : f = x^3+y^2*z+y*z^2;
i3 : frobeniusPower( 1/2, ideal f )
o3 = ideal (x, y, z)
o3 : Ideal of R
i4 : u=f^3;
i5 : frobeniusPower( 1/4, ideal u )
o6 = ideal (x, y, z)
o6 : Ideal of R
i7 : ((HSLGModule(1, f))#3) + 1
o7 : 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/11[x,y,z];
i2 : f = y^2*z^3-x^5-2*z^5;
i3 : frobeniusPower( 1/11, ideal f^10 )
             2        3
o6 = ideal (z , x*z, x )
o3 : Ideal of R
i5 : frobeniusPower( 1/121, ideal f^120 )
             2        3
o6 = ideal (z , x*z, x )
o6 : Ideal of R
i7 : ((HSLGModule(1, f))#3) + 1
o7 : 2
\end{MyVerbatim}
}\medskip

>>>>

Again, consider the maps $T^e : \omega_{R^{1/p^e}} \to \omega_R$ we have considered throughout this section.  It is a theorem of Hartshorne-Speiser, Lyubeznik, and Gabber \cite{HartshorneSpeiserLocalCohomologyInCharacteristicP,LyubeznikFModulesApplicationsToLocalCohomology,Gabber.tStruc} that the descending images
\begin{equation}\label{eq: descending images}
\omega_R \supseteq T(\omega_{R^{1/p}}) \supseteq \dots \supseteq T^e(\omega_{R^{1/p^e}}) \supseteq T^{e+1}(\omega_{R^{1/p^{e+1}}}) \supseteq \cdots
\end{equation}
stabilize for $e \gg 0$.
The function \texttt{FPureModule} computes the stable submodule in this chain, called the \emph{HSLG module}, and returns a list containing the following items:
\begin{enumerate}[(1)]
   \item The HSLG module;
   \item The canonical module, embedded (non-uniquely) as an ideal;
   \item The element $u$ representing the map on the canonical module (see
   \autoref{subsec.DualToFrobeniusOnQuotientRings});
   \item The value of $e > 0$ at which the descending images in \eqref{eq: descending images} stabilize.  This is sometimes also called the HSLG-number of the canonical module as a Cartier module.
\end{enumerate}

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/3[x,y,z]/ideal( x^3 + y^4 + z^5 );
i2 : L = FPureModule R;
i3 : L#0
                          2        2   3
o3 = ideal (0, y*z, x*z, y , x*y, x , z )
o3 : Ideal of R
i4 : L#1
o4 = ideal 1
o4 : Ideal of R
i5 : L#3
o5 = 1
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/2[x,y,z];
i2 : f = x^3+y^2*z+y*z^2;
i3 : frobeniusPower( 1/2, ideal f )
o3 = ideal (x, y, z)
o3 : Ideal of R
i4 : u=f^3;
i5 : frobeniusPower( 1/4, ideal u )
o6 = ideal (x, y, z)
o6 : Ideal of R
i7 : ((FPureModule(1, f))#3) + 1
o7 : 2
\end{MyVerbatim}
}\medskip

@@

\medskip
{\small
\begin{MyVerbatim}
i1 : R = ZZ/11[x,y,z];
i2 : f = y^2*z^3-x^5-2*z^5;
i3 : frobeniusPower( 1/11, ideal f^10 )
             2        3
o6 = ideal (z , x*z, x )
o3 : Ideal of R
i5 : frobeniusPower( 1/121, ideal f^120 )
             2        3
o6 = ideal (z , x*z, x )
o6 : Ideal of R
i7 : ((FPureModule(1, f))#3) + 1
o7 : 2
\end{MyVerbatim}
}\medskip

### 2019-04-24 Fsing/TestIdealsPaper/TestIdeals_REMOTE_5127.tex
Renamed HSLGModule to FPureSubmodule.

We can extend the definition of Frobenius powers as follows
\begin{definition}[cf. \cite{HernandezTeixeiraWittFrobeniusPowers}]
Let  $I\subseteq S$ be an ideal.
\begin{enumerate}
 \item[(a)] If $a$ is a positive integer with base-$p$ expansion  $a=a_0 + a_1 p +  \dots + a_r p^r$, we define
 $I^{[n]}=I^{a_0} \left(I^{a_1}\right)^{[p]} \dots  (I^{a_r})^{[p^r]}$. %\left( I^{a_r}\right)^{[p^r]}$ .
 \item[(b)] If $t$ is a non-negative rational number of the form $t = a/p^e$, we define  $I^{[t]} = (I^{[a]})^{[1/p^e]}.$
 \item[(c)] If $t$ is any non-negative rational number, and $\{a_n/p^{e_n}\}_{n\geq 1}$ is a sequence of rational numbers converging to $t$ from above, we define $I^{[t]}$
 to be the stable value of the increasing chain of ideals $\{I^{[a_n/p^{e_n}]}\}_{n\geq 1}$.

@@

The following two examples describe two prototypical
$p^{-e}$- and $p^{e}$-linear maps.
\begin{example}
For any $S$-module $M$, we can construct a new $S$-module $M^{1/p^e}$ with elements $\{  m^{1/p^e} \,|\,m\in M\}$ by defining
$m_1^{1/p^e} +  m_2^{1/p^e} =  (m_1 +  m_2)^{1/p^e}$ for all $m_1, m_2 \in M$ and
$s  m^{1/p^e}=  (s^{p^e} m)^{1/p^e}$ for all $m\in M$ and $s\in S$.

Consider any $\phi\in \Hom_S(M^{1/p^e}, M)$: if we identify $M^{1/p^e}$ with $M$ we can interpret $\phi$ as a $p^{-e}$-linear map.
\end{example}

@@

\subsection{Implementation and complexity}

The computation of Frobenius roots $(-)^{[1/p^e]}$ is the computational workhorse behind many of the methods in \emph{TestIdeals}
hence it is important to understand how it is computed and its computational complexity.

Let $R=\mathbb{K}[x_1, \dots, x_n]$ and fix $B$ to be $\mathbb{K}$-basis for $\mathbb{K}^{1/p^e}$;
it is not hard to see that $R^{1/p^e}$ is a free $R$-module with free basis
$$\left\{ b x_1^{\alpha_1/p^e} \dots x_n^{\alpha_n/p^e} \,|\, 0\leq \alpha_1, \dots, \alpha_n < p^e \right\}.$$

\begin{proposition}(cf. \cite[\S 5]{KatzmanParameterTestIdealOfCMRings})
\begin{enumerate}
\item[(a)] If the ideal $I\subseteq R$ is generated by $g_1, \dots, g_k\in R$, then $I^{[1/p^e]}=\sum_{i=1}^k (g_i R)^{[1/p^e]}$.
\item[(b)] If
\begin{equation}\label{eqn1}
g=\bigoplus_{b\in B, \atop{0\leq \alpha_1, \dots, \alpha_n < p^e}} r_{b, \alpha_1, \dots, \alpha_n} b x_1^{\alpha_1/p^e} \dots x_n^{\alpha_n/p^e},
\end{equation}
then $(gR)^{[1/p^e]}$ is generated by $\{ r_{b, \alpha_1, \dots, \alpha_n} \,|\, b\in B, 0\leq \alpha_1, \dots, \alpha_n < p^e \}$.
\end{enumerate}
\end{proposition}

Thus the computation of Frobenius roots of ideals reduces to the case of principal ideals, and the complexity grows linearly with the number of generators of the ideal.
Furthermore the calculation of  $(gR)^{[1/p^e]}$ reduces to finding the summands in (\ref{eqn1}), which essentially amounts
taking each term in $g$, computing the $p^e$th root of the coefficient in $\mathbb{K}$, and dividing the monomial exponent vector by $p^e$ with remainder.
Thus the complexity of  computing $(gR)^{[1/p^e]}$ is proportional to the number of terms in $g$ and is independent of its degree.
\emph{The calculation of Frobenius roots does not involve the calculation of Gr\"obner bases.}

@@

In this section, we explain how to compute parameter test modules, parameter test ideals, test ideals and $F$-Pure-modules\footnote{$F$-pure-modules (i.e. HSLG-modules) can be used to give a scheme structure to the $F$-injective or $F$-pure locus.}.

@@

\subsection{FPureSubmodule, computing $F$-pure submodules of rank-1 Cartier modules}

@@

Recall that in section \ref{Section: p-linear maps} we identified $p^{-e}$ linear maps
with elements of $\Hom_R( R^{1/p^e}, R)$ and it turns out that this is a principal $R^{1/p^e}$-module generated
the \emph{trace map} $\pi\in \Hom_R(^{1/p^e}, R)$, constructed as follows (cf. \cite[Lemma 1.6]{FedderFPureRat} and \cite[Example 1.3.1]{BrionKumarFrobeniusSplitting}).
Recall that if $B$ is a $\mathbb{K}$-basis for $\mathbb{K}^{1/p^e}$;
$R^{1/p^e}$ is a free $R$-module with basis
$$\left\{ b x_1^{\alpha_1/p^e} \dots x_n^{\alpha_n/p^e} \,|\, 0\leq \alpha_1, \dots, \alpha_n < p^e \right\} ;$$
the trace map $\pi$ is the projection onto the free summand
$R x_1^{(p^e-1)/p^e} \dots x_n^{(p^e-1)/p^e}\cong R$.

We can now write our given $\phi$ as multiplication by some $u^{1/p^e}$ followed by $\pi$ and it is not hard to see that
an ideal $I\subseteq R$ is $\phi$-compatible if and only if $u I \subseteq I^{[p^e]}$.

@@

The defining condition $u I = I^{[p]}$ for $\phi$-compatible ideals allows us to
think of these in a dual form: write $\mathfrak{m}=(x_1, \dots, x_n)R$,
$E=E_{R\mathfrak{m}}(R_{\mathfrak{m}}/\mathfrak{m})=\HH^n_{\mathfrak{m}} (R)$, and let $T: E \rightarrow E$ be the $p^e$-linear map
induced from the Frobenius map on $R$.
If $\psi=u T$, then $\psi \Ann_E I \subseteq \Ann_E I$ if and only if $u I = I^{[p]}$!
Thus finding all $R$-submodules of $E$ compatible with $\psi=u T$ also amounts to finding all
$\phi=\pi \circ u^{1/p^e}$-compatible ideals.

>>>>

We can extend the definition of Frobenius powers as follows
\begin{definition}[cf. \cite{HernandezTeixeiraWittFrobeniusPowers}]
Let  $I\subseteq S$ be an ideal.
\begin{enumerate}
 \item[(a)] If $a$ is a positive integer with base-$p$ expansion  $a=a_0 + a_1 p +  \dots + a_r p^r$, we define
 $I^{[n]}=I^{a_0} \left(I^{a_1}\right)^{[p]} \dots  (I^{a_r})^{[p^r]}$. %\left( I^{a_r}\right)^{[p^r]}$ .
 \item[(b)] If $t$ is a non-negative rational number of the form $t = a/p^e$, we define  $I^{[t]} = (I^{[a]})^{[1/p^e]}.$
 \item[(c)] If $t$ is any non-negative rational number, and $\{a_n/p^{e_n}\}_{n\geq 1}$ is a sequence of rational numbers converging to $t$ from above, we define $I^{[t]}$
 to be the stable value of increasing chain of ideals $\{I^{[a_n/p^{e_n}]}\}_{n\geq 1}$.

@@

The following two examples describe two prototypical
$p^{-e}$- and $p^{e}$-linear maps.
\begin{example}
For any $S$-module $M$, we can construct a new $S$-module $F_*^e M$ with elements $\{ F_*^e m \,|\,m\in M\}$ by defining
$F_*^e m_1 + F_*^e m_2 = F_*^e (m_1 +  m_2)$ for all $m_1, m_2 \in M$ and
$s F_*^e m= F_*^e s^{p^e}$ for all $m\in M$ and $s\in S$.

Consider any $\phi\in \Hom_S(F_*^e M, M)$: if we identify $F_*^e M$ with $M$ we can interpret $\phi$ as a $p^{-e}$-linear map.
\end{example}

@@

@@

In this section, we explain how to compute parameter test modules, parameter test ideals, test ideals and HSLG-modules\footnote{HSLG-modules can be used to give a scheme structure to the $F$-injective or $F$-pure locus.}.

@@

\subsection{HSLG module, computing $F$-pure submodules of rank-1 Cartier modules}

@@

Recall that in section \ref{Section: p-linear maps} we identified $p^{-e}$ linear maps
with elements of $\Hom_R(F_*^e R, R)$ and it turns out that this is a principal $F_*^e R$-module generated
the \emph{trace map} $\pi\in \Hom_R(F_*^e R, R)$, constructed as follows (cf. \cite[Lemma 1.6]{FedderFPureRat} and \cite[Example 1.3.1]{BrionKumarFrobeniusSplitting}).
Let $B$ be a $\mathbb{K}^{p^e}$-basis for $\mathbb{K}$;
$F_*^e R$ is a free $R$-module with basis
$$\left\{ F_*^e b x_1^{\alpha_1} \dots x_n^{\alpha_n} \,|\, 0\leq \alpha_1, \dots, \alpha_n < p^e \right\}$$
and the trace map $\pi$ is the projection onto the free summand
$R F_*^e  x_1^{p^e-1} \dots x_n^{p^e-1}\cong R$.

We can now write our given $\phi$ as multiplication by some $F_*^e u$ followed by $\pi$ and it is not hard to see that
an ideal $I\subseteq R$ is $\phi$-compatible if and only if $u I \subseteq I^{[p^e]}$.

@@

The defining condition $u I = I^{[p]}$ for $\phi$-compatible ideals allows us to
think of these in a dual form: write $\mathfrak{m}=(x_1, \dots, x_n)R$,
$E=E_{R\mathfrak{m}}(R_{\mathfrak{m}}/\mathfrak{m})=\HH^n_{\mathfrak{m}} (R)$, and let $T: E \rightarrow E$ be the $p^e$-linear map
induced from the Frobenius map on $R$.
If $\psi=u T$, then $\psi \Ann_E I \subseteq \Ann_E I$ if and only if $u I = I^{[p]}$!
Thus finding all $R$-submodules of $E$ compatible with $\psi=u T$ also amounts to finding all
$\phi=\pi \circ F_*^e u$ ideals.
